C Code Conversion to Counter Automata for Reachability Analysis

### 2020-03-12 Documentation/paper.tex
updated paper

@@

\title{Reachability problems for counter automata}
\author{Lars Van Roy\\
\textit{dept. of Mathematics and Computer Science} \\
\textit{University of Antwerp}\\
lars.vanroy@student.uantwerpen.be}
\date{\today}

@@

\section{Introduction}
A well known problem in writing code is the reachability of specific lines in set written code. It is often impossible to properly determine whether or not a line of code guarded by a condition statement is going to be reachable in any possible execution of the code. Unreachable code, due to unsatisfiable conditions, is an obvious issue for time restrained environments such as embedded computer systems. In order to upgrade the efficiency of code that is affected by time constraints, or code in general, a proper reachability analysis would be of great use. \newpage

In order to analyze code reachability, we will make use of a property proven by Daniel Bundala and Joal Ouaknine \cite{danialandjoel}, in which they state that reachability is decidable for timed automata with a single parametric clock. Adding the restriction of only allowing a single parametric clock has a big effect on the types of code we can analyse, but is unavoidable, since there is no proof for two counter machines. Three counter machines and up have been proven to be undecidable.

In this paper we will give an overview of a program which will be able to convert c code to such an automaton. The program will go through the following iterations on the path to generating a counter automaton.

\begin{enumerate}
	\item convert c code into a parse tree, using antlr
	\item perform context free reductions on this parse tree
	\item perform context sensitive reductions on this parse tree
	\item validate the resulting parse tree, to be conforming to the code constraints
	\item generate the counter automaton
\end{enumerate}

The c code can at most contain one counter (and preferably exactly one, as zero counter programs are assumed to be trivial when considering reachability). Furthermore, the operations on the counter will be restricted, the counter can only be altered via assignments with parameters or constants, and the counter can only be compared using equality, inequality, strict less than, strict greater than, less than or equal, greater than or equal, and only to paremeters or constants. Furthermore, these parameters can never be altered throughout the duration of the code.

@@

\section{One counter automata}
The automata used in this project are a specific subset of the generally defined automata. The general definition for an automaton A is given by the tuple (S, $s_0$, $\delta$, F, Op, $\lambda$) where
\begin{itemize}
	\item S is the set of states
	\item $s_0$ is the set of initial states
	\item $\delta$ is the set of edges
	\item F is the set of final states
	\item Op is the set of operations
	\item $\lambda$ is the mapping for every edge to an element of the set of operations
\end{itemize}

A class of automata derived from the general automaton is the class of timed automata. Before we can consider a definition for these automata, we need to introduce a few additional concepts as described Martin Fr\^anzle et al. \cite{informationprocessingletters} and Uli Fahrenberg \cite{higherdimensionaltimedautomata}.

Consider a set of clocks $\chi$, the corresponding set $\phi(\chi)=\{x_1,...,x_n\}$ of clock constraints is generated corresponding to the following grammar $\varphi ::= true | x<k | x = k | x>k | \varphi\land\varphi$. where $k \in \mathbb{N}$ and $x \in \chi$. Hence $\phi(\chi)$ is nothing but a set of constraints based on conjunctions of comparisons with clocks to integers.

A clock valuation is a mapping $v:\chi \rightarrow \mathbb{R}_{\geq 0}$ where $\mathbb{R}_{\geq 0}$ is the set of non-negative real numbers. We denote $\mathbb{R}^{\chi}_{\geq 0}$ as the set of all clock valuations. Furthermore we write $v \models \varphi$ to denote that $v \in \mathbb{R}^{\chi}_{\geq 0}$ satisfies the constraint $\varphi$. We denote by 0 the valuation such that 0(x) = 0 for all $x \in \chi$. Given $t \in \mathbb{R}^{\chi}_{\geq 0}$, we let v+t be the clock valuation such that (v+t)(x) = v(x) + t for all clocks $x \in \chi$. Given $\lambda \subseteq \chi$, let v[$\lambda \leftarrow 0$] be the clock valuation such that
\[
	v[\lambda \leftarrow 0](x) = 
	\begin{cases}
		0,& \text{if } x \in \lambda\\
		v(x),& \text{if } x \notin \lambda
	\end{cases}
\]
A timed automaton B is defined by a tuple B = $(Q, q^0, Q^f, I, E)$, where
\begin{itemize}
	\item $Q$ is a finite set of locations
	\item $q^0$ represents the initial location, $q^0 \in Q$
	\item $Q^f$ represents the set of accepting locations, $Q^f \subseteq Q$
	\item $I$ assigns invariants to states, $I: Q \rightarrow \phi(\chi)$
	\item $E$ is the set of guarded transitions, $E \subseteq Q$ x $\phi(\chi)$ x $\Sigma$ x $2^\chi$ x $Q$
\end{itemize}

A run in a timed automaton is a sequence $\pi$ of configurations $c_1c_2 ... c_k$ such that each configuration is a direct successor of the previous configuration or, in other words, $c_i \rightarrow c_{i+1}$. We call a run accepting \textsl{accepting} if for a run $c_1c_2...c_k$ it holds that $c_1$ is the initial state and $c_k$ is the final state.

A one-counter automaton is a tuple $\mathcal{A}$ = $(Q, q_0, \sigma, F)$ where
\begin{itemize}
	\item $Q$ is a finite set of states
	\item $q_0 \in Q$ is the initial state
	\item $F \subseteq Q$ is the set of accepting states
	\item $\sigma \subseteq Q $ x $ L $ x $ Q $ is the transition relation with L the set of instructions
\end{itemize}

A counter valuation v is an element of $\mathbb{N}$ and a configuration of $\mathcal{A}$ is a pair in Q x $\mathbb{N}$. The initial configuration is the pair ($q_i$, 0).

Finally, we will consider the following set of allowed operations where P is the set of parameters, the $\wedge$ symbol will be used to represent an empty label.
\[
 op = \{+c, -c, +p, -p, \leq c, =c, \geq c, \leq p, =p, \geq p: c \in \mathbb{N}, p \in P\} \cup \{\wedge\}
\]

The automata we will consider will be one-counter automata. The set $Q$ will contain a state for each line of code. The initial state $q_0$ will be the state corresponding with the first line of code. For now the set of final states $F$ will be empty, but the intended definition of the final state is the line(s) of which the reachability will be checked. The transition relation will consist of elements of the form $(q_i, l, q_{i+1})$ where $q_i, q_{i+1} \in Q$ and $l \in op$. For each element it will hold that we can reach line $q_{i+1}$ from $q_i$. The l element or operation can either be blank, in which case there is a statement on line $q_i$ but it has no effect on the evaluation of our counter and will therefore be disregarded, or not blank. The operation will always be part of the \textit{op} set described earlier.

There are a few exception to the above described states, some segments will have additional nodes to avoid confusion. In a loop there can be a pre loop, which will signify the state after the initial \textit{for condition} is performed, a start node for the inner segment, an end node to the inner segment, a post loop node, that will signify the state after the post statement has been evaluated, and finally a \textit{post iteration node}, which will signify the end of the statements. Same holds for selection statements (eg. if statements) which will have a start node, a start of success node, which signifies the body of the if statement, a start of else node, which will signify possible if else or else statements, and an end of selection node. Other states can be added for statements that require multiple steps, like assignments etc. These will be further explained in the generator section.
\subsection{division example}
Consider the following code segment, this code snippet will attempt to determine whether or not a given integer a, is greater or equal compared to a given integer b. Throughout the paper, the different steps will be described which will ultimately yield the automaton displayed in figure \ref{fig:greater_or_equal}.

\begin{lstlisting}[style=CStyle]
bool is_greater_or_equal(int a, int b){
	if(a >= b){
		return true;
	}
	return false;
}
\end{lstlisting}

\begin{figure}[h]
	\includegraphics[width=\linewidth]{is_greater_or_equal_counter_automaton}
	\caption{A counter automaton corresponding with the is\_greater\_or\_equal function.}
	\label{fig:greater_or_equal}
\end{figure}

This example automaton $\mathcal{A}$ is defined by the tuple $(Q, q_0, \sigma, F)$, where
\begin{align*}
	Q &= \{1, 1.97, 1.98, 1.99, 2, 2.01, 2.02, 2.03, 2.04, 3, 5\} \\
	q_0 &= 1 \\
	\sigma &= \left\{ \begin{array}{l}
		(1, \wedge, 1.97), (1.97, \text{\textless =a}, 1.98), (1.98, \text{+1}, 1.98), (1.98, \text{=a}, 2), \\
		(1.97, \text{\textgreater =a}, 1.99), (1.99, \text{-1}, 1.99), (1.99, \text{=a}, 2), (2, \text{\textgreater =b}, 2.01),\\ (2.01, \wedge, 3), (2, \text{+1}, 2.03), (2.03, \text{\textless =b}, 2.04), (2.04, \text{-1}, 2.02),\\ (2.04, \wedge, 5)
	\end{array} \right\}\\
	F &= \emptyset
\end{align*}

\section{Program}
\subsection{Program overview}
The program that will perform the conversion from c code to a one counter automaton, will do this in five steps.

@@

We will allow all counter operations that can directly be supported by the $op$ set described earlier. We will add assignment and inequality to this, as these can easily be represented as sequences of operations in the $op$ list. Along with these we will also need support for strict less than and greater than, which can also be implemented using the operations in the \textit{op} list. The full explanation on how this can be done, can be found in the generator section.

\subsection{Initialization}
The program will start of by running the generator antlr compilation unit on the given c code. In case there are any issues regarding simple compilation rules, the program will exit with an error, indicating the issues in the code. For the remainder of the execution, the code is assumed to be correct, and there will be no regard for potential issues within the code.

\subsection{Context-free abstract syntax tree generation}
Using the parse tree resulting from the antlr compilation, we will try to generate an abstract syntax tree that is as simple as possible without regarding any context.

First of all, the code uses the notion of a node stack, this node stack will be used to store the nodes that are currently being evaluated.

The first loop will operate as a visitor, which will traverse the parse tree in a depth first manner. The first discovery of a node is done by calling the enter function corresponding to the type of node. As soon as all children of set node have been evaluated, the corresponding exit function will be called.

@@

\subsubsection{Context-free analysis assignmentExpression example}
To look at a concrete example, consider the following rule in the grammar:
\begin{align*}
	&assignmentExpression\\
	&\text{:}\indent \text{conditionalExpression}\\
	&|\indent \text{unaryExpression assignmentOperator assignmentExpression}\\
	&|\indent \text{Digitsequence}; \\
\end{align*}
The rule for conditionalExpression needs to be there, as there is a chain of expressions, and some expressions need to be higher in the order than others, therefore all expressions with a higher priority need to be evaluated before the lower priority expressions. This is done by adding a rule to a lower priority expression in each expression class. It is however undesirable for these nodes to be added. Without regard for context in other nodes, we can simply look at the current node to see whether or not the conditionalExpression line is used (which antlr generated context allows us to do).

@@

\subsection{Context-sensitive abstract syntax tree reduction}
Now that all obviously redundant nodes have been eliminated, there is still the need for context sensitive reductions. We will attempt to implement constant propagation and substitution, in which we compute constant expressions, and replace variables with known value, by their actual value. Furthermore we will try to rephrase expressions by expressions with the same meaning, so that all expressions, if possible, are in an acceptable format for the counter automaton.

@@

\subsubsection{Context-sensitive reduction example constant propagation}
Another example of a needed reduction, is constant propagation and folding. It is possible that a code segment would initially be rejected, due to use of unwanted variables, while in essence, these variables are nothing but variable representations of results of constant expressions. Consider the code snippet below. It would initially be rejected, because the counter variable is initialized with a variable that is neither a constant nor a parameter.
\newpage
\begin{lstlisting}[style=CStyle]
bool divide(int p, int n){
	int variable = 5;
	variable += 20;
	int counter = variable;
	while(counter > 0){
		counter -= n;
	}
	counter += 1;
	counter -= 1;
	if(counter == 0){
		return true;
	}
	return false;
}
\end{lstlisting}

@@

\subsubsection{Context-sensitive reduction example addition folding}
As described earlier, we do allow additions and subtractions of the form - v and + v, where $v \in P$ or $v \in \mathbb{N}$ with $P$ the set of parameters. Given that the formation described in figure \ref{fig:unfolded_addition} occurs in the automaton, we would technically not be allowed to accept this, as this is not directly in the specified format.

@@

However, the context-sensitive reduction will be able to reduce the earlier specified example to the automaton in figure \ref{fig:folded_addition} This automaton will be acceptable, as this is a direct use of an allowed operation.

@@

Whenever such a node occurs, the current situation will be inspected. We can either be assigning to a counter, or in a restricted conditional expression. The correct error will be chosen if either of the two situations is relative.

@@

\subsubsection{validation example}
When we go back to the earlier mentioned segment of code. We obtain the automaton displayed in figure \ref{fig:divide_cleaned}.

@@

Next thing we encounter is the Iteration statement, for which we will check the condition with the constrained conditional variable enabled. With this variable enabled, we can simply continue traversing the nodes, and in case there are any unsupported nodes, the corresponding error would automatically be generated. However, there will be no corresponding nodes, so we will leave the Relational Expression without any error statuses, and we will disable the constrained conditional variable.

@@

The selection statement will be evaluated in the exact same manner as the Iteration Statement was evaluated, and as the jump statement holds no special operations, we can conclude that this function is ok for counter automaton generation.

\subsection{Counter automaton generation}
The final section of the program will be the counter generation itself. In this loop, there will be no more regards to possible invalid statements. At this point we assume everything to be known, and we can therefore simplify things drastically. Recall that the following list is the collection of allowed labels for the counter automaton.
\[
op = \{+c, -c, +p, -p, \leq c, =c, \geq c, \leq p, =p, \geq p: c \in \mathbb{N}, p \in P\} \cup \{\wedge\}
\]
This allows us to just check the head nodes related to such expressions.

An important thing to remark, is that we are not interested in general properties of runs for these automata. We are interested in the possibility of reachability or in other words, the existence of one path that will take us to a given node. This assumption is later on used to justify certain compositions.

@@

\subsubsection{Assignments}
Other than Functions, we will also need to know when assignments to counters occur, so that we can add the proper transition labels. We only need to check whether or not we are assigning to a counter, if so, we read the operation used, and the variable that gets assigned, and the manner $\in \{=, +=, -=\}$ gets used.

As mentioned earlier, assignments are not in the set of operations, but they can be represented using the operators displayed earlier, as can be seen in figure \ref{fig:counter_assignment}. This can be done by using a reset operation, where we first check whether or not the counter is smaller or equal to the desired value, if so, it is either already satisfied, in which case we can just carry on, or it is smaller than the desired value, for which we will add a self loop with $+1$ as a label, so that we will eventually reach the desired value. The same sequence of events occur in case the counter is larger or equal to the desired value, with the exception that we will use $-1$ as an in between operation.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.48\linewidth]{counter_assignment}
	\caption{Assignment of a value p to a counter.}
	\label{fig:counter_assignment}
\end{figure}

@@

This can be done without adding additional weightings to transitions, due to the remark made earlier. We are not interested in properties for every single run, yes there are runs where the counter can loop infinitely, but there will also be a run where the counter gets to the correct value, after which it will continue onward.

\subsubsection{Loop statements}
We will also need the notion of selection statements. These will make use of a maximum of 5 nodes to enforce everything related to the loops.

The first node will be the $start$ of the conditional statement, the second node is the pre node. This node symbolizes the state after the precondition of the for loop has been evaluated.

Next we need a node to identify the $start\ of\ the\ inner\ segment$, which will be reached by a transition with the condition as a label.

After this node the inner segment will follow, and it will be finalized with the $stop\ inner\ segment$ node, in case this one is needed. This stop node will symbolize the state after the post expression of the for loop has been evaluated, after this node, a transition will occur to the pre node, to reevaluate the loop condition.

@@

Consider the following segment of code.
\begin{lstlisting}[style=CStyle]
bool test(int a){
	int counter = 25;
	for(counter = 0; counter < 5; counter ++){
		continue;
	}
	counter = counter + 5;
	return true;
}
\end{lstlisting}

\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{test_counter_automaton}
	\caption{Example of an Iteration Statement.}
	\label{fig:test_counter_automaton}
\end{figure}

This will result in the counter automaton seen in figure \ref{fig:test_counter_automaton}. The iteration start node will be node 3.01, the start of the inner segment will be node 3.06 and the end of this same segment 3.02.

The inner segment itself consists out of one single node, being node 4.

The end of the iteration is denoted by node 3.03, and the first node beyond the iteration statement, being node 6, will follow after this one.

\subsubsection{Switch statements}
A second conditional expression that exists in c, are \textit{switch statements}. These statements allow a variable to be compared against multiple values. What is special about the switch statements, is that if matched with one case, the code of all consequential cases will also be executed, until the end of the switch, or until a jump occurs.

@@

\begin{lstlisting}[style=CStyle]
bool test(int a){
	switch(a){
		case 5:
			printf("The parameter is greater than 4.");
		case 3:
			printf("The parameter is greater than 2.");
		default:
			printf("The parameter value is %d", a);
	}

	return true;
}
\end{lstlisting}

To generate a counter automaton for a switch statement, we will start by generating a chain for the statements in the cases. This is represented by the chain starting from node 4, all the way to node 8, in figure \ref{fig:counter_switch}.

After that we need to add the conditional branches that correspond to the different cases. A first branch originates from node 2, and can either go to node 2.02 in case the counter is equal to 5, or to 2.08 in case the counter is not.

There is a second branch from 2.08 where we can either go to 2.04 in case the counter is equal to 3, or to 2.06 in case the counter is not.

The case starting points are represented by 2.02, 2.04 and 2.06 respectively, and their respective code follows directly beyond these nodes.

\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{counter_switch}
	\caption{Example of a switch statement.}
	\label{fig:counter_switch}
\end{figure}
\subsubsection{If statements}
Another important construct are the if statements. These statements will, just as the loop statements did, require additional nodes to be able to be modeled. If statements require a total of 4 nodes to be modeled.

First of all, we need a $start\ of\ the\ iteration\ statement$ node, from which the conditions will start.

We will need an $if\ node$ and an $else\ node$ which will both signify the start of the if and possible else segment.

After the inner segment, they will all end up at the $end\ of\ the\ iteration\ statement$ from which point they will continue with the next line of code.

Consider the test2 function, it contains an if statement, an else if statement and an else statement.
\begin{lstlisting}[style=CStyle]
bool test2(){
	int counter = 25;
	if(counter == 23){
		counter += 3;
	}
	else if(counter == 24){
		counter += 2;
	}
	else{
		counter += 1;
	}
	return true;
}
\end{lstlisting}

@@

The program will return a counter automaton as displayed in figure \ref{fig:test2_counter_automaton}. The iteration start node of the outer if will be node 3, the if node and else node of the same if will be 3.01 and 3.04 respectively. The end node of this if statement will be 3.02.

The else if will have 6 as a start node, 6.01 and 6.04 will function as if and else nodes respectively. Node 6.02 will signify the end of the else if statement.

Finally, the else statement will start at node 7, since this is no conditional statement, it will just behave as regular code.

\subsubsection{Inequality condtions}
Another statement we will need to support, are inequality conditions. If we want our conditional statements to be properly functioning, we will need to be able to generate opposing conditions, but the opposing condition of the equality condition, is the inequality condition, which we can not directly model using the given operations.

The following code segment makes use of an if statement which has an equality condition in it. This means that the edge going towards the else node, will have the opposing condition as a label.

\begin{lstlisting}[style=CStyle]
bool is_ten(int a){
	int counter = a;

	if(counter == 10){
		return true;
	}

@@

When looking at part of the resulting counter automaton in figure \ref{fig:inequality_expression}, we will notice the transitions going from the start of the if statement, being node 4, towards the end of the if statement, being node 4.02. There will not be an else node, as there is no else if or else statement.

@@

The first transition chain starts with \textit{+1} followed by \textit{\textless =10}, followed by \textit{-1}, which results in the same expression as saying the original counter is strictly smaller than 10. The final \textit{-1} is there to reset the counter, so that the counter is back to what it originally was.

>>>>

\usepackage{nameref}

\usepackage{lscape}

@@

\title{C Code Conversion to Counter Automata for Reachability Analysis}
\author{Lars Van Roy\\
\textit{dept. of Mathematics and Computer Science} \\
\textit{University of Antwerp}\\
lars.vanroy@student.uantwerpen.be}
\date{\today}

@@

\section{Introduction}
A well known problem in writing code is the reachability of specific lines in said written code. It is often impossible to properly determine whether or not a line of code guarded by a conditional statement is going to be reachable in any possible execution of the code\cite{713323, 10.1145/1292316.1292319}. Unreachable code is never desired, but is an ever more pressing issue for energy restrained environments such as embedded computer systems \cite{embeddedefficiency}. In these systems, we want our execution to be as efficient as possible, and therefore avoid conditions that can never be satisfied or lines that can never be reached at any cost. In order to upgrade the efficiency of code that is affected by time constraints, or code in general, a proper reachability analysis would be of great use.

In order to analyse code reachability, we will want to make use of one counter automata. The one counter automata are especially interesting as code is easily convertible to counter automata and it has been proven \cite{succintandparametriconecounter, danialandjoel} that reachability is decidable for this type of automata. As will be shown in the continuation of this paper, adding the restriction of only allowing a single parametric clock has a big effect on the types of code we can analyse, but is unavoidable, since there is no proof for two counter machines. Reachability problems for machines with three or more counters have been proven to be undecidable.

To analyse code reachability in (a subset of) c programs, we propose a program that will go through the following iterations on the path to generate a counter automaton.

\begin{enumerate}
	\item convert c code into a parse tree, using ANTLR
	\item perform context free reductions on this parse tree
	\item perform context sensitive reductions on this parse tree
	\item validate the resulting parse tree, to be conforming to the code constraints
	\item generate the counter automaton
\end{enumerate}

We will only consider a subset of c programs, which contain at most one counter where a counter is a variable that is used inside a conditional statement. The program will be able to handle c code with zero counters, but this is considered trivial and will be disregarded for the remainder of this paper. Furthermore, the operations on the counter will be restricted, the counter can only be updated via assignments with parameters or constants, and the counter can only be compared using equality, inequality, strict less than, strict greater than, less than or equal, greater than or equal, and only to parameters or constants. Furthermore, the function parameters are not allowed to be altered throughout the duration of the code.

@@

\section{Motivation}
Popular IDE's do already implement reachability analysis up to a certain degree. However, two of the most used IDE's, being Eclipse and Clion, were tested, and they both show that in some code snippets, reachability analysis is not performed up until the degree that could be done using counter automata. It is capable to resolve trivial dead code (aka constant conditions or code post return etc), but it is not able to resolve impossible sequences, where no possible counter value would ever satisfy some of the conditions in the sequence.

This proves that there is a definite need for additional research on analysis in code. Reachability analysis by using counter automata would not only allow for an expansion of the current reachability analysis as is, but might also allow further expansions based on the results of the counter automaton expansion.

\section{Preliminaries}
This paper will make use of counter automata and more specifically one counter automata. These automata are derived from the general definition of automata and signify an extension by adding the notion of counters.

\subsection{General automata}
The automata used in this research are a specific subset of the generally defined automata. The general definition for an automaton A is given by the tuple (Q, $\Sigma$, $\delta$, $q_0$, F) where
\begin{itemize}
	\item Q is the set of states
	\item $\Sigma$ is a finite set of symbols, called the alphabet of the automaton
	\item $\delta$ is the transition function, $\delta: Q \times \Sigma \rightarrow Q$
	\item $q_0$ is the initial state, $q_0 \in Q$
	\item F is a set of states of Q defining the final states, $F \subseteq Q$
\end{itemize}

\subsection{Counter automata}

A one-counter automaton is a tuple $\mathcal{A}$ = $(Q, q_0, \sigma, op, F)$ where
\begin{itemize}
	\item $Q$ is a finite set of states
	\item $q_0 \in Q$ is the initial state
	\item $F \subseteq Q$ is the set of accepting states
	\item $op$ is the set of operations
	\item $\sigma \subseteq Q \times op \times Q$ is the transition relation with $op$ the set of operations
\end{itemize}

The counter valuation $c \in \mathbb{N}$ is the value that will be analysed, updated during the evaluation of the automaton.

At all times the automaton will be in a configuration $(q, c)$ where $q \in Q$ and $c$ the counter valuation where $(q_0, 0)$ is the initial configuration.

Finally, we will consider the following set of allowed operations where $P$ is the set of parameters, the $\epsilon$ symbol will be used to represent an empty label.
\[
 op = \{+c, -c, +p, -p, \leq c, =c, \geq c, \leq p, =p, \geq p: c \in \mathbb{N}, p \in P\} \cup \{\epsilon\}
\]

The automata we will consider will be one-counter automata. The set $Q$ will contain a state for each line of code. The initial state $q_0$ will be the state corresponding with the first line of code. For now the set of final states $F$ will be empty, but the intended interpretation of a final state will be the line of which the reachability will be checked. The transition relation will consist of elements of the form $(q_i, l, q_{i+1})$ where $q_i, q_{i+1} \in Q$ and $l \in op$.

There are a few exceptions to the above described states, some segments will have additional nodes to avoid confusion. These special nodes, and their meaning, will be further explained in their relevant sections inside Section \ref{sec:generation}, \nameref{sec:generation}.

\section{Program}
The program that will perform the conversion from c code to a one counter automaton, will do this in five steps.

@@

We will allow all counter operations that can directly be supported by the $op$ set described earlier. We will add assignment and inequality to this, as these can easily be represented as sequences of operations in the $op$ list. Along with these we will also need support for strict less than and greater than, which can also be implemented using the operations in the \textit{op} list. The full explanation on how this can be done, can be found in Section \ref{sec:generation}.

\subsection{Initialization}
The program will start of by running the generator ANTLR compilation unit on the given c code. In case there are any issues regarding simple compilation rules, the program will exit with an error, indicating the issues in the code. For the remainder of the execution, the code is assumed to be correct, and there will be no regard for potential issues within the code.

\subsection{Context-free abstract syntax tree generation}
Using the parse tree resulting from the ANTLR compilation, we will try to generate an abstract syntax tree that is as simple as possible without regarding any context.

First of all, the code uses the notion of a node stack. This node stack will be used to store the nodes that are currently being evaluated.

The first loop will operate as a visitor, which will traverse the parse tree in a depth first manner. The first discovery of a node is done by calling the enter function corresponding to the type of node. As soon as all children of said node have been evaluated, the corresponding exit function will be called.

@@

\subsubsection{Context-free analysis assignmentExpression example}
To look at a concrete example, consider the following rule in the grammar:
\begin{align*}
	&assignmentExpression\\
	&\text{:}\indent \text{conditionalExpression}\\
	&|\indent \text{unaryExpression assignmentOperator assignmentExpression}\\
	&|\indent \text{Digitsequence}; \\
\end{align*}
The rule for conditionalExpression needs to be there, as there is a chain of expressions, and some expressions need to be higher in the order than others, therefore all expressions with a higher priority need to be evaluated before the lower priority expressions. This is done by adding a rule to a lower priority expression in each expression class. It is however undesirable for these nodes to be added. Without regard for context in other nodes, we can simply look at the current node to see whether or not the conditionalExpression line is used (which ANTLR generated context allows us to do).

@@

\subsection{Context-sensitive abstract syntax tree reduction}
There is still the need for context sensitive reductions. We will attempt to implement constant propagation and substitution, in which we compute constant expressions, and replace variables with known value, by their actual value. Furthermore we will try to rephrase expressions by expressions with the same meaning, so that all expressions, if possible, are in an acceptable format for the counter automaton.

@@

\subsubsection{Context-sensitive reduction example: constant propagation}
Another example of a needed reduction, is constant propagation and folding. It is possible that a code segment would initially be rejected, due to use of unwanted variables, while in essence, these variables are nothing but variable representations of results of constant expressions. Consider the code snippet below. It would initially be rejected, because the counter variable is initialized with a variable that is neither a constant nor a parameter.
\newpage
\begin{lstlisting}[style=CStyle]
bool divide(int p, int n){
	int variable = 5;
	variable += 20;
	int counter = variable;
	while(counter > 0){
		counter -= n;
	}
	counter += 1;
	counter -= 1;
	if(counter == 0){
		return true;
	}
	return false;
}
\end{lstlisting}

@@

\subsubsection{Context-sensitive reduction example: addition folding}
As described earlier, we do allow additions and subtractions of the form $-x$ and $+x$, where $x \in P$ or $x \in \mathbb{N}$ with $P$ the set of parameters. Given that the formation described in Figure \ref{fig:unfolded_addition} occurs in the automaton, we would technically not be allowed to accept this, as this is not directly in the specified format.

@@

However, the context-sensitive reduction will be able to reduce the earlier specified example to the automaton in Figure \ref{fig:folded_addition}. This automaton will be acceptable, as this is a direct use of an allowed operation.

@@

Whenever we encounter a node part of the $unsupported$ set, the current situation will be inspected. For example, if a \textit{Multiplication Expression} node occurs, inside a subtree of an \textit{Assignment Expression} node where an assignment to a counter was specified, we know that we are in an invalid state, as counters can only be updated via assignments, additions or subtractions.

The same logic holds for subtrees of conditional expressions. In case we have a conditional expression where the subtree contains a \textit{Bitwise And Expression} node, we know that this is an invalid state, as constant conditions have already been folded in the previous step, this must mean that there is a counter in the condition, but counter comparisons are not allowed to contain and statements.

@@

\subsubsection{validation example}
When we go back to the earlier mentioned segment of code. We obtain the automaton displayed in Figure \ref{fig:divide_cleaned}.

@@

Next, we encounter the Iteration statement, for which we will check the condition with the constrained conditional variable enabled. With this variable enabled, we can simply continue traversing the nodes, and in case there are any unsupported nodes, the corresponding error would automatically be generated. However, there will be no corresponding nodes, so we will leave the Relational Expression without any error statuses, and we will disable the constrained conditional variable.

@@

The selection statement will be evaluated in the exact same manner as the Iteration Statement was evaluated, and as the jump statement holds no special operations, we can conclude that this function satisfies the requirements for one counter automaton generation.

\subsection{Counter automaton generation}
\label{sec:generation}
The final section of the program will be the counter generation itself. In this loop, there will be no more regards to possible invalid statements. At this point we assume everything to be known, and we can therefore simplify things drastically. Recall that the following list is the collection of allowed labels for the counter automaton.
\[
op = \{+c, -c, +p, -p, \leq c, =c, \geq c, \leq p, =p, \geq p: c \in \mathbb{N}, p \in P\} \cup \{\wedge\}
\]
This allows us to just check the head nodes related to such expressions.

@@

\subsubsection{Assignments}
Other than functions, we will also need to know when assignments to counters occur, so that we can add the proper transition labels. We only need to check whether or not we are assigning to a counter, if so, we read the operation used, and the variable that gets assigned. Valid counter assignments need to be of the form \textit{c op x} where $c$ is the counter, $op \in $ \textit{\{=, +=, -=\}} and $x \in P$ or $x \in \mathbb{N}$.

As mentioned earlier, assignments are not in the set of operations, but they can be represented using the operators displayed earlier, as can be seen in Figure \ref{fig:counter_assignment}. This can be done by using a reset operation, where we first check whether or not the counter is smaller or equal to the desired value. If so, it is either already satisfied, in which case we can just carry on, or it is smaller than the desired value, for which we will add a self loop with $+1$ as a label, so that we will eventually reach the desired value. 
\begin{figure}[h]
	\centering
	\includegraphics[width=0.48\linewidth]{counter_assignment}
	\caption{Assignment of a value p to a counter.}
	\label{fig:counter_assignment}
\end{figure}

@@

An important thing to remark, is that we are not interested in general properties of runs for these automata. We are interested in the possibility of reachability or in other words, the existence of one path that will take us to a given node. In the given example, there is a path where we would remain in node 3.98 indefinitely but since there is at least one path that would take us to node 4, node 4 is still considered reachable.

\subsubsection{Iteration statements}
We will also need the notion of iteration statements, which needs 5 additional special nodes to fully support it's functionality. The first node will be the \textit{start of the loop}. The second node is the \textit{pre node}, this node symbolizes the state after the precondition of the for loop has been evaluated. Next we need a node to identify the \textit{start of the inner segment}, which will be reached by a transition with the condition as a label.

After this node the inner segment will follow, and it will be finalized with the \textit{stop inner segment} node, in case this one is needed. This stop node will symbolize the state after the post expression of the for loop has been evaluated, after this node, a transition will occur to the pre node, to re-evaluate the loop condition.

@@

Consider the following segment of code, this will result in the automaton in Figure \ref{fig:test_counter_automaton}.
\begin{lstlisting}[style=CStyle]
bool test(int c){
	for(c = 0; c < 5; c ++){
		continue;
	}
	return true;
}
\end{lstlisting}

\begin{figure}[h]
	\centering
	\includegraphics[width=0.95\linewidth]{test_counter_automaton}
	\caption{Example of an Iteration Statement.}
	\label{fig:test_counter_automaton}
\end{figure}

\subsubsection{Switch statements}
A second conditional statement that exists in c, is a \textit{switch statement}. This statement allows a variable to be compared against multiple values. What is special about the switch statements, is that if matched with one case, the code of all consequential cases will also be executed, until the end of the switch, or until a jump occurs.

@@

\begin{lstlisting}[style=CStyle]
bool test(int a){
	switch(a){
		case 3:
			printf("The parameter is greater than 2.");
		case 2:
			printf("The parameter is greater than 1.");
		default:
			printf("The parameter value is %d", a);
	}
	return true;
}
\end{lstlisting}

To generate a counter automaton for a switch statement, we will start by generating a chain for the statements in the cases. This is represented by the chain starting from \textit{start\_switch}, all the way to \textit{default}, in Figure \ref{fig:counter_switch}.

After that we need to add the conditional branches that correspond to the different cases. A first branch originates from \textit{start\_switch}, and can either go to \textit{case\_3} in case the counter is equal to 3, or to \textit{not\_case\_3} in case the counter is not.

There is a second branch from \textit{not\_case\_3} where we can either go to \textit{case\_2} in case the counter is equal to 2, or to \textit{not\_case\_2} in case the counter is not. Finally, from \textit{not\_case\_2} we will always go to the \textit{default} node.

\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{counter_switch}
	\caption{Example of a switch statement.}
	\label{fig:counter_switch}
\end{figure}
\subsubsection{If statements}
\label{sec:if}
Another important construct are the if statements. These statements will, just as the iteration statements did, require additional nodes to be able to be modelled. If statements require a total of 4 nodes to be modelled.

First of all, we need a $start\ if$ node, from which the conditions will start.

We will need an $if$ node and an $else$ node which will both signify the start of the if and possible else segment.

After the inner segment, they will all end up at the $end\ if$ node from which point they will continue with the next line of code.

Consider the test2 function, and the corresponding simplified counter automaton.
\begin{lstlisting}[style=CStyle]
bool test2(int c){
	if(c == 1){
		printf("The counter is equal to 1.");
	}
	else if(c == 2){
		printf("The counter is equal to 2.");
	}
	return true;
}
\end{lstlisting}

@@

The program will return a more in depth version of the simplified counter automaton displayed in Figure \ref{fig:test2_counter_automaton}. The first if statement will start from $start\_if\_1$, and end at $end\_if\_1$, the else if will start from $start\_if\_2$, and end at $end\_if\_2$.

\subsubsection{Inequality conditions}
\label{sec:inequality}
Another statement we will need to support, are inequality conditions. If we want our conditional statements to be properly functioning, we will need to be able to generate opposing conditions, but the opposing condition of the equality condition, is the inequality condition, which we can not directly model using the given operations.

The following code segment makes use of an if statement which has an equality condition in it, being the edge going towards the else node.

\begin{lstlisting}[style=CStyle]
bool is_ten(int c){
	if(c == 10){
		return true;
	}

@@

When looking at part of a simplified counter automaton in Figure \ref{fig:inequality_expression}, we will notice the transitions going from the \textit{start\_if} node, towards the \textit{end\_if} node. There will not be an \textit{else} node, as there is no \textit{else if} or \textit{else} statement.

@@

The first transition chain follows the \textit{lte} nodes, which starts with \textit{c += 1} followed by \textit{c \textless =10}, followed by \textit{c -= 1}, which results in the same expression as saying the original counter is strictly smaller than 10. The final \textit{c -= 1} is there to reset the counter, so that the counter is back to what it originally was.
